%% 摘要

% 中文摘要
\begin{abstract}
大语言模型后训练的关键，是将人类偏好或专家行为转化为可优化的奖励信号。无论显式训练奖励模型，还是通过直接偏好优化等方式隐式刻画偏好目标，其核心都在于为策略更新提供偏好导向的优化信号。现有方法在这一过程中仍面临多重问题：偏好数据和人类标注可能引入主观偏差；奖励信号可能随着策略分布变化而发生分布失配；若缺少对奖励更新幅度的限制，奖励尺度和语义容易漂移，进而诱发奖励过优化和训练不稳定。

针对上述问题，本文研究大语言模型偏好对齐中的奖励函数学习，目标是从专家示范数据中恢复奖励信号，并在策略分布变化时保持稳定。本文提出近端逆奖励优化方法PIRO（Proximal Inverse Reward Optimization），以专家示范响应作为偏好行为证据，在策略 - 奖励迭代更新的逆强化学习框架下直接从专家示范数据中学习奖励函数，减少对额外成对偏好比较数据的依赖，缓解偏好比较环节带来的标注噪声影响；同时，PIRO限制相邻轮次奖励变化，使奖励函数适应当前策略分布并避免无约束漂移。理论上，本文将对齐过程建模为奖励学习与策略优化耦合的双层问题，推导局部替代目标相对于原始对齐目标的改进下界，说明奖励变化误差项会影响原始目标的稳定改进。算法实现上，本文以相邻轮次奖励输出的均方根差（RMSD）近似奖励变化，并用自适应约束系数形成可监测、可自适应调节的稳定性反馈。

本文在TL;DR摘要生成和UltraFeedback通用指令跟随任务上进行验证。TL;DR任务中，PIRO在第五轮迭代中取得奖励评分$1.415$，较固定奖励RLHF/PPO提升约$14.1\%$，较SFT提升约$30.0\%$；去除奖励变化约束后，模型早期短暂上升，随后持续退化，说明该约束对维持多轮 IRL 训练稳定性具有关键作用。UltraFeedback任务中，PIRO最佳检查点MT-Bench Avg(8)为$7.53$，超过SFT、DPO、RLHF(PPO)、RLHF(GRPO)以及无约束IRL基线，并在推理和STEM等维度上取得提升。实验结果表明，PIRO 在动态奖励学习过程中能够同时提升策略性能与奖励稳定性。

\keywordslist{大语言模型对齐, 逆强化学习, 奖励模型, RMSD约束, 策略优化}
\end{abstract}

% 英文摘要
\begin{engabstract}
Aligning large language models in post-training comes down to turning human preferences or expert behavior into an optimizable reward signal. Whether a method trains an explicit reward model or shapes preference targets implicitly through direct preference optimization, the reward signal is what drives policy updates toward preferred behavior. Existing approaches still face several difficulties: preference data and human annotation may introduce subjective bias; the reward signal can become misaligned with the policy as the policy keeps changing; and without limits on how much the reward model is allowed to change, its scale and semantics drift, inviting reward overoptimization and unstable training.

To address these problems, this thesis studies reward-function learning for preference alignment of large language models, aiming to recover a reward signal from expert demonstrations and keep it stable as the policy distribution shifts. The thesis proposes Proximal Inverse Reward Optimization (PIRO). Treating expert demonstration responses as evidence of preferred behavior, PIRO learns the reward function directly within an inverse reinforcement learning framework that alternates between policy and reward updates, reducing reliance on additional pairwise preference labels and the annotation noise they carry; PIRO also bounds reward changes between adjacent iterations, so the reward model can adapt to the current policy without drifting unconstrained. On the theoretical side, alignment is formulated as a bilevel problem coupling reward learning and policy optimization, and an improvement lower bound for a local surrogate objective relative to the original alignment objective is derived, showing how the reward-change error term affects stable improvement of the original objective. On the algorithmic side, the root mean square difference (RMSD) between reward outputs across adjacent iterations approximates the reward change, and an adaptive coefficient turns this into an observable, self-adjusting stability signal.

The method is evaluated on TL;DR summarization with Pythia-1B and UltraFeedback instruction following with Qwen3.5-9B. On TL;DR, PIRO reaches a reward score of $1.415 \pm 0.054$ at iter5, $14.1\%$ above RLHF/PPO and $30.0\%$ above SFT; removing the reward-change constraint yields a brief early gain followed by sustained degradation, indicating that this constraint is essential for keeping multi-round IRL training stable. On UltraFeedback, PIRO's best checkpoint scores $7.53$ on MT-Bench Avg(8), ahead of SFT, DPO, RLHF(PPO), RLHF(GRPO), and the unconstrained IRL baseline, with clearer gains on reasoning and STEM. The results indicate that PIRO improves both policy performance and reward stability under dynamic reward learning.

\engkeywordslist{LLM alignment, IRL, reward model, RMSD constraint, policy optimization}
\end{engabstract}
